Do we need detailed demographic data to forecast plant population responses to climate change?

### 2016-10-28 manuscript/manuscript.Rmd
small updates to ms correction

Despite the weak climate effects, including climate covariates did increase the \textcolor{blue}{mean} accuracy of forecasts for two species: *B. gracilis* and *Poa secunda* (\textcolor{blue}{Fig. S3}).
\textcolor{blue}{However, skill increases were only statistically significant at $\alpha$ = 0.05 for the QBM for \emph{B. gracilis} QBM ($t_{(279)}$ = 1.80, \emph{P} = 0.037).}
Forecast error decreased significantly with the inclusion of climate covariates for the *B. gracilis* IPM ($t_{(280)}$ = -3.72, *P* = \textcolor{blue}{0.01}) and QBM ($t_{(280)}$ = -3.34, *P* < 0.0001), and for the *Poa secunda* IPM ($t_{(196)}$ = -1.90, *P* textcolor{blue}{= 0.015}) and QBM ($t_{(196)}$ = -2.47, *P* = 0.007) (Fig. S1, Supporting Information).
In no case did including climate covariates significantly decrease forecast skill (Table S21), despite small changes in the mean skill (Fig. 2, \textcolor{blue}{Fig. S3}).

@@

The higher accuracy of the IPM and QBM with climate covariates for *B. gracilis* and *Poa secunda*  highlights the advantage of contemporary modeling and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
Ridge regression allows researchers to retain covariates whose effects may be difficult to identify in noisy data or short time series.
This is especially important when forecasting the impacts of climate variability, where it is important to include the effects of forcing variables (e.g., temperature and precipitation) even if such effects are difficult to identify.
Indeed, we failed to detect strong climate effects in the QBM for *B. gracilis* and *Poa secunda*, but including climate covariates still improved forecasting skill (Fig. 2).
If a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate toward zero [@Hooten2015].
Of course, regardless of what model selection approach is adopted, a critical step is identifying the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
However, the climate covariates we chose required aggregating daily weather data over discrete time periods.
It is possible that we did not choose the optimal time periods over which to aggregate.
New methods using functional linear models (or splines) may offer a data-driven approach for identifying the appropriate time periods over which to aggregate to produce a tractable set of candidate climate variables [@Sims2007; @VandePol2011; @Teller2016].

@@

Acknowledgments
---------------
This work was funded by the National Science Foundation through a Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370), award MSB-1241856 to MBH, and a CAREER award to PBA (DEB-1054040). We thank the original mappers of the permanent quadrats in Montana and the digitizers in the Adler lab, without whom this work would not have been possible. Informal conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a series of meetings between the Adler and Weecology labs at USU sharpened our thinking. Brittany Teller provided comments that improved our manuscript. Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. government. This research was supported by the Utah Agricultural Experiment Station, Utah State University, and approved as journal paper number 8917.

>>>>

Despite the weak climate effects, including climate covariates did increase the \textcolor{blue}{mean} accuracy of forecasts for two species: *B. gracilis* and *Poa secunda* (\textcolor{blue}{Fig. S3}).
\textcolor{blue}{However, skill increases were only statistically significant at $\alpha$ = 0.05 for the QBM for \emph{B. gracilis} QBM ($t_{(279)}$ = 1.80, \emph{P} = 0.037).}
Forecast error decreased significantly with the inclusion of climate covariates for the *B. gracilis* IPM ($t_{(280)}$ = -3.72, *P* = \textcolor{blue}{0.01}) and QBM ($t_{(280)}$ = -3.34, *P* < 0.0001), and for the *Poa secunda* IPM ($t_{(196)}$ = -1.90, *P* \textcolor{blue}{= 0.015}) and QBM ($t_{(196)}$ = -2.47, *P* = 0.007) (Fig. S1, Supporting Information).
In no case did including climate covariates significantly decrease forecast skill (Table S21), despite small changes in the mean skill (Fig. 2, \textcolor{blue}{Fig. S3}).

@@

The higher accuracy \textcolor{blue}{and lower error} of the IPM and QBM with climate covariates for *B. gracilis* and *Poa secunda*  highlights the advantage of contemporary modeling and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
Ridge regression allows researchers to retain covariates whose effects may be difficult to identify in noisy data or short time series.
This is especially important when forecasting the impacts of climate variability, where it is important to include the effects of forcing variables (e.g., temperature and precipitation) even if such effects are difficult to identify.
Indeed, we failed to detect strong climate effects in the QBM for *B. gracilis* and *Poa secunda*, but including climate covariates still improved forecasting skill (Fig. 2, \textcolor{blue}{Fig. S3}) \textcolor{blue}{and reduced forecasting error (Fig. S1)}.
If a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate toward zero [@Hooten2015].
Of course, regardless of what model selection approach is adopted, a critical step is identifying the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
However, the climate covariates we chose required aggregating daily weather data over discrete time periods.
It is possible that we did not choose the optimal time periods over which to aggregate.
New methods using functional linear models (or splines) may offer a data-driven approach for identifying the appropriate time periods over which to aggregate to produce a tractable set of candidate climate variables [@Sims2007; @VandePol2011; @Teller2016].

@@

Acknowledgments
---------------
This work was funded by the National Science Foundation through a Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370), award MSB-1241856 to MBH, and a CAREER award to PBA (DEB-1054040). We thank the original mappers of the permanent quadrats in Montana and the digitizers in the Adler lab, without whom this work would not have been possible. Informal conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a series of meetings between the Adler and Weecology labs at USU sharpened our thinking. Brittany Teller provided comments that improved our manuscript. Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. government. This research was supported by the Utah Agricultural Experiment Station, Utah State University, and approved as journal paper number 8917. \textcolor{blue}{And, thanks to Andy Kleinhesselink for discovering an error in our computer code before this paper was published.}

### 2016-10-28 manuscript/manuscript.tex
small updates to ms correction

Despite the weak climate effects, including climate covariates did
increase the \textcolor{blue}{mean} accuracy of forecasts for two
species: \emph{B. gracilis} and \emph{Poa secunda}
(\textcolor{blue}{Fig. S3}).
\textcolor{blue}{However, skill increases were only statistically significant at $\alpha$ = 0.05 for the QBM for \emph{B. gracilis} QBM ($t_{(279)}$ = 1.80, \emph{P} = 0.037).}
Forecast error decreased significantly with the inclusion of climate
covariates for the \emph{B. gracilis} IPM (\(t_{(280)}\) = -3.72,
\emph{P} = \textcolor{blue}{0.01}) and QBM (\(t_{(280)}\) = -3.34,
\emph{P} \textless{} 0.0001), and for the \emph{Poa secunda} IPM
(\(t_{(196)}\) = -1.90, \emph{P} textcolor\{blue\}\{= 0.015\}) and QBM
(\(t_{(196)}\) = -2.47, \emph{P} = 0.007) (Fig. S1, Supporting
Information). In no case did including climate covariates significantly
decrease forecast skill (Table S21), despite small changes in the mean
skill (Fig. 2, \textcolor{blue}{Fig. S3}).

@@

The higher accuracy of the IPM and QBM with climate covariates for
\emph{B. gracilis} and \emph{Poa secunda} highlights the advantage of
contemporary modeling and variable selection approaches such as ridge
regression and LASSO over techniques that would exclude
``non-significant'' effects from final models. Ridge regression allows
researchers to retain covariates whose effects may be difficult to
identify in noisy data or short time series. This is especially
important when forecasting the impacts of climate variability, where it
is important to include the effects of forcing variables (e.g.,
temperature and precipitation) even if such effects are difficult to
identify. Indeed, we failed to detect strong climate effects in the QBM
for \emph{B. gracilis} and \emph{Poa secunda}, but including climate
covariates still improved forecasting skill (Fig. 2). If a species is
truly unresponsive to a given climate variable, statistical
regularization techniques will shrink the mean and variance of a
covariate estimate toward zero (Hooten and Hobbs 2015). Of course,
regardless of what model selection approach is adopted, a critical step
is identifying the appropriate candidate covariates, which we attempted
to do based on our knowledge of this semi-arid plant community. However,
the climate covariates we chose required aggregating daily weather data
over discrete time periods. It is possible that we did not choose the
optimal time periods over which to aggregate. New methods using
functional linear models (or splines) may offer a data-driven approach
for identifying the appropriate time periods over which to aggregate to
produce a tractable set of candidate climate variables (Sims et al.
2007, Pol and Cockburn 2011, Teller et al. 2016).

@@

This work was funded by the National Science Foundation through a
Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370), award
MSB-1241856 to MBH, and a CAREER award to PBA (DEB-1054040). We thank
the original mappers of the permanent quadrats in Montana and the
digitizers in the Adler lab, without whom this work would not have been
possible. Informal conversations with Stephen Ellner, Giles Hooker,
Robin Snyder, and a series of meetings between the Adler and Weecology
labs at USU sharpened our thinking. Brittany Teller provided comments
that improved our manuscript. Compute, storage and other resources from
the Division of Research Computing in the Office of Research and
Graduate Studies at Utah State University are gratefully acknowledged.
Any use of trade, firm, or product names is for descriptive purposes
only and does not imply endorsement by the U.S. government. This
research was supported by the Utah Agricultural Experiment Station, Utah
State University, and approved as journal paper number 8917.

>>>>

Despite the weak climate effects, including climate covariates did
increase the \textcolor{blue}{mean} accuracy of forecasts for two
species: \emph{B. gracilis} and \emph{Poa secunda}
(\textcolor{blue}{Fig. S3}).
\textcolor{blue}{However, skill increases were only statistically significant at $\alpha$ = 0.05 for the QBM for \emph{B. gracilis} QBM ($t_{(279)}$ = 1.80, \emph{P} = 0.037).}
Forecast error decreased significantly with the inclusion of climate
covariates for the \emph{B. gracilis} IPM (\(t_{(280)}\) = -3.72,
\emph{P} = \textcolor{blue}{0.01}) and QBM (\(t_{(280)}\) = -3.34,
\emph{P} \textless{} 0.0001), and for the \emph{Poa secunda} IPM
(\(t_{(196)}\) = -1.90, \emph{P} \textcolor{blue}{= 0.015}) and QBM
(\(t_{(196)}\) = -2.47, \emph{P} = 0.007) (Fig. S1, Supporting
Information). In no case did including climate covariates significantly
decrease forecast skill (Table S21), despite small changes in the mean
skill (Fig. 2, \textcolor{blue}{Fig. S3}).

@@

The higher accuracy \textcolor{blue}{and lower error} of the IPM and QBM
with climate covariates for \emph{B. gracilis} and \emph{Poa secunda}
highlights the advantage of contemporary modeling and variable selection
approaches such as ridge regression and LASSO over techniques that would
exclude ``non-significant'' effects from final models. Ridge regression
allows researchers to retain covariates whose effects may be difficult
to identify in noisy data or short time series. This is especially
important when forecasting the impacts of climate variability, where it
is important to include the effects of forcing variables (e.g.,
temperature and precipitation) even if such effects are difficult to
identify. Indeed, we failed to detect strong climate effects in the QBM
for \emph{B. gracilis} and \emph{Poa secunda}, but including climate
covariates still improved forecasting skill (Fig. 2,
\textcolor{blue}{Fig. S3})
\textcolor{blue}{and reduced forecasting error (Fig. S1)}. If a species
is truly unresponsive to a given climate variable, statistical
regularization techniques will shrink the mean and variance of a
covariate estimate toward zero (Hooten and Hobbs 2015). Of course,
regardless of what model selection approach is adopted, a critical step
is identifying the appropriate candidate covariates, which we attempted
to do based on our knowledge of this semi-arid plant community. However,
the climate covariates we chose required aggregating daily weather data
over discrete time periods. It is possible that we did not choose the
optimal time periods over which to aggregate. New methods using
functional linear models (or splines) may offer a data-driven approach
for identifying the appropriate time periods over which to aggregate to
produce a tractable set of candidate climate variables (Sims et al.
2007, Pol and Cockburn 2011, Teller et al. 2016).

@@

This work was funded by the National Science Foundation through a
Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370), award
MSB-1241856 to MBH, and a CAREER award to PBA (DEB-1054040). We thank
the original mappers of the permanent quadrats in Montana and the
digitizers in the Adler lab, without whom this work would not have been
possible. Informal conversations with Stephen Ellner, Giles Hooker,
Robin Snyder, and a series of meetings between the Adler and Weecology
labs at USU sharpened our thinking. Brittany Teller provided comments
that improved our manuscript. Compute, storage and other resources from
the Division of Research Computing in the Office of Research and
Graduate Studies at Utah State University are gratefully acknowledged.
Any use of trade, firm, or product names is for descriptive purposes
only and does not imply endorsement by the U.S. government. This
research was supported by the Utah Agricultural Experiment Station, Utah
State University, and approved as journal paper number 8917.
\textcolor{blue}{And, thanks to Andy Kleinhesselink for discovering an error in our computer code before this paper was published.}
